---
title: "Diel.Niche.Vignette"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Diel.Niche Vignette}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{bayesplot,ggplot2,coda,Diel.Niche}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

#### Authors: Brian Gerber & Mason Fidino
#### Date: 2023-06-15

# Introduction

The $\texttt{Diel.Niche}$ package is used to evaluate how animals use the three fundamental periods of light: twilight (dawn & dusk), daytime, and nighttime. First, install and load the library

```{r package, echo = FALSE}
# Install package from GitHub via devtools.
if(
  !"Diel.Niche" %in% row.names(installed.packages())
){
  devtools::install_github(
    "diel-project/Diel-Niche-Modeling",
    build_vignettes = TRUE
  )
}

# Load the package
library(Diel.Niche)
```

The packages specifies many diel hypotheses, which together are defined into six hypothesis sets. The code names for the hypotheses and sets are,

```{r hyp}
what.hyp()
```

A short description of each hypothesis within a given set is available by calling
it within the `what.hyp()` function. For example, `"CR"` represents the 
`General Crepuscular` hypothesis within the `Traditional` hypothesis set.

```{r hyp2}
  what.hyp("CR")
```

To list all of the hypotheses of a given set you can use `hyp.sets()`

```{r hyp3}
  hyp.sets("Traditional")

  hyp.sets("General")
```

The best way to understand each hypothesis set is by plotting them. Of the six 
currently available hypothesis sets within `Diel.Niche`, we consider four of them as different fundamental ways to investigate how a species uses diel time, which includes the Maximizing, Traditional, General, and Selection sets. We
have left out the Variation and Threshold hypothesis sets from this vignette, but if you are interested in those you can read about themn in the `Diel-Niche-Additional` vignette.

## **Maximizing**

This hypothesis set includes three hypotheses [Crepuscular Max (`CR.Max`), Diurnal Max (`D.Max`), and Nocturnal Max (`N.Max`)] with the objective to evaluate which time period is used most. As such, there is no hypothesis about activity across multiple time periods (i.e., cathemeral). To plot this, or any of the currently available hypothesis sets within `Diel.Niche`, you can use the `triplot()` function.


```{r plotting 3d,fig.align = "center",fig.height = 6, fig.width = 6,out.width = "5in"}
triplot(
  hyp = hyp.sets(
    "Maximizing"
  )
)
```

Because `Diel.Niche()` quantifies how a species uses three separate diel periods (twilight, day, and night), `triplot()` plots the proportion of time a species may use each period on a three dimensional plot with the `plotly()` package. If you use `RStudio` and are having issues seeing plots from `triplot()`, go to
the bottom of this vignette.

Hovering over the figure generated via `triplot()` shows the proportion of time
a species could use each diel period to possibly be classified under a given hypothesis. For example, if the proportion of time a species uses daytime (the y axis) is > 0.5, then the Diurnal Max (`D.max`) would likely be the most supported hypothesis (depending on sample size).

## **Traditional**


This hypothesis set includes four hypotheses [Crepuscular (`CR`), Diurnal (`D`), Nocturnal (`N`), Traditional Cathemeral (`C`)] that aim to capture the general interpretation of these hypotheses from the literature. Crepuscular, Diurnal, and Nocturnal are defined based on having at least 0.80 probability (threshold probability, $\xi_{1} = 0.80)$ in their respective diel periods (twilight, daytime, nighttime). If an animal is not mostly active in one period than it is defined as Traditional Cathemeral; this occurs when either two or three time periods are used more than $1-\xi_{1}$. The logic behind the threshold of 0.80 is that an animal uses a clear majority of time in one diel period, but is not so strict that there is not some moderate amount of activity outside of this period. While this is the default value we set for the traditional hypothesis set in $\texttt{Diel.Niche}$, it can easily be modified by a user when fitting these hypotheses to their data (see below).

```{r plotting3d 2,fig.align = "center",fig.height = 6, fig.width = 6,out.width = "5in"}
triplot(
  hyp = hyp.sets(
    "Traditional"
  )
)
```


## **General** 

This hypothesis set includes seven hypotheses. The Diurnal (`D`), Crepuscular (`CR`), and Nocturnal (`N`) hypotheses are defined the same as in Traditional. The main difference is the separation of the probability space of Traditional Cathemeral into four more specific hypotheses: General Cathemeral (`C2`), Crepuscular-Nocturnal (`CR.N`), Diurnal-Nocturnal (`D.N`), and Diurnal-Crepuscular (`D.CR`). The General Cathemeral hypothesis (`C2`)---which represents a subset of the parameter space taken by the previously mentioned Traditional Cathemeral (`C`)---aims to  define when an animal uses all three diel periods (twilight, daytime, nighttime) at equal to or more than a minimum amount (i.e., $p_{\text{tw}}, p_{\text{d}}, p_{\text{n}} \leq \xi_{1}$  and $p_{\text{tw}}, p_{\text{d}}, p_{\text{n}} \geq \xi_{2}$). We defined the lower threshold probability as $\xi_{2} = 0.10$, such that we consider it important to differentiate animal activity when a diel period is used at least this much. However, if only two diel periods are used above $\xi_{2}$ then we classify this activity using one of the binomial hypotheses (Crepuscular-Nocturnal, Diurnal-Nocturnal, and Diurnal-Crepuscular). 

```{r plotting3dGeneral 2,fig.align = "center",fig.height = 6, fig.width = 6,out.width = "5in"}
triplot(
  hyp = hyp.sets(
    "General"
  )
)
```

For example, suppose a species is active mainly during the day with $p_{\text{d}} = 0.78$, relatively active during twilight with $p_{\text{tw}} = 0.16$, and not very active at night with $p_{\text{n}} = 0.06$. We would define this activity as Diurnal-Crepuscular because $p_{\text{tw}}$ \& $p_{\text{d}} \geq \xi_{2}$, while $p_{\text{n}} < \xi_{2}$. However, if night activity was also higher than $\xi_{2}$, such that $p_{\text{d}} = 0.70$, and $p_{\text{tw}} = p_{\text{n}} = 0.15$ then we classify this activity as General Cathemeral because a moderate to large amount of activity occurs in all three diel periods. In summary, the Traditional hypothesis set distinguishes between unimodal and multi-modal diel activity, while the General set distinguishes among unimodal, bimodal, and trimodal activity. 

**Selection** 

This hypothesis set includes seven diel phenotype hypotheses and are defined based on an inputted amount of proportional time available to an animal in each diel period ($\mathbf{p}_{\text{avail}} = [p_{av.tw},\;p_{av.d}]$). The available time in the night period does not need to be input as it is derived as $1 - p_{av.tw}, - p_{av.d}]$. 

The hypotheses in this set are Day Selection (`D.avail`), Twilight Selection (`TW.avail`), Night Selection (`N.avail`), use equivalent to availability (`EQ.avail`), Day-Twilight Selection (`D.TW.avail`), Night-Twilight Selection (`N.TW.avail`), and Day-Night Selection (`D.N.avail`). Unlike the other hypothesis sets that aim to quantify the dominant time periods a species uses, the selection hypothesis set quantifies if a species uses a given diel period more than it is available. 

One specific component that is important to consider with this hypothesis set is that the amount of time available in each diel period changes based on the day of the year and location, and can be calculated with relative ease with the `suncalc` package in R.

First, let's plot where day and night have equal availability as, 
```{r 3dplot7,fig.align = "center",fig.height = 6, fig.width = 6,out.width = "6.5in"}
triplot(
  hyp = hyp.sets(
    "Selection"
  ),
  diel.setup = diel.ineq(
    p.avail = c(0.04,0.48)
  )
)
```

We can see the hypotheses change when we make an increasing amount of available time in the night period, such as during winter in the higher latitude areas,

```{r 3dplot8,fig.align = "center",fig.height = 6, fig.width = 6,out.width = "6.5in"}
triplot(
  hyp = hyp.sets(
    "Selection"
  ),
  diel.setup = diel.ineq(
    p.avail = c(0.04,0.3)
  )
)
```

Because of the shift in availability throughout a year, researchers may want to 
discretize longer time periods into smaller pieces to determine if selection varies, for example, over the course of a year.

# Plotting a single hypothesis

You can also use `triplot` to also view specific hypotheses. For example, to 
plot out the Traditional diurnal hypothesis you would run

```{r 3dplot6,fig.align = "center",fig.height = 6, fig.width = 6,out.width = "6.5in"}
triplot(
  hyp = "D",
  more.points = TRUE
)
```

Note that while this looks like a full triangle, the axes are quite different (
i.e., this is zoomed in on the part of the diel parameter space that is associated to this hypothesis within the traditional hypothesis set
). Because this is a zoomed in look at one hypothesis, we set the `more.points` argument to `TRUE` in order to better fill in the space to view it.

# Simulating Data

To simulate data based on a diel hypothesis, we need define a hypothesis using its coded name. Pick a hypothesis to simulate data from and how many samples.

```{r sim1}
  set.seed(45451)
  hyp <- c("CR")
  sim <- sim.diel(
    hyp = hyp,
    n.sample = 100
  )
  
  #The probability value used to simulate data
  (p <- sim$p)
  
  #The simulated data
  (y <- sim$y)

```

# Setup hypothesis set and fit models

You can setup your own model set by creating a vector of coded hypothesis names or use the pre-defined model sets. In general, we suggest using all of the hypothesis within a given set instead of a handful of them, but `Diel.Niche` can be used both ways.

```{r hyp.set}
#Manual 
  hyp.set <- c("D","CR") 
  
#Pre-defined
  hyp.set <- hyp.sets("Traditional")
```

To fit the data to the hypothesis set using default and equal priors for each model set, all you need to do is supply the data and which hypothesis to fit to `diel.fit()`. Furthermore, the Bayes factors table (`bf.table`) can be accessed to check out the posterior probability of support for each hypothesis given the data. 

```{r fit}
  out <- diel.fit(
    y,
    hyp.set = hyp.sets(
      "Traditional"
    )
  )

#Call the model probabilities for each hypothesis in the set
  out$bf.table
```
In this case, the support for `CR` is almost 1, which makes sense as we simulated data under that hypothesis for this example.

The returned diel list object has numerous named elements, all of which are detailed in the help file for `diel.fit`

```{r results, eval = FALSE}
attributes(out)

?diel.fit
```
```{r show_results, echo = FALSE}
attributes(out)
```

The function `diel.fit` defaults to providing the model probabilities for each hypothesis set, but not the posterior samples of the parameters for each hypothesis. If you would like the posterior MCMC samples, which can be useful for plotting purposes and model reporting, you can change the `post.fit` argument to `TRUE`. Futthermore, if you would like to assess model convergence then multiple chains are necessary. This can be modified by chaning the `n.chains` argument

```{r results2}
  out <-  diel.fit(
    y,
    hyp.set = hyp.sets("Traditional"),
    n.chains = 2,
    post.fit = TRUE
  )
```

We can look at a convergence criteria for each hypothesis. Note that convergence can be very poor for models that do not meet the inequality constraints well. As, such, we highly suggest running these models on multiple chains and ensuring that the best fit model has converged.
```{r convergence}
  out$gelm.diag
```

In this example, the Nocturnal model had issues with estimating `p_crep_1` (Probability of crespuscular actvity). More specifically, the Nocturnal model estimated it as zero across both chains, making it impossible to compute a gelman diagnostic. Regardless, this 'lack of convergence' as specified by this diagnostic should have little effect on the calculation of the Bayes factors, so there is no need to remove the Nocturnal model from this set. What this does mean, however, is that inference should not be made from the Nocturnal model. 

We can plot the posterior samples from the most supported model to check convergence/mixing as well. You can access the posterior of the best fit model via `post.samp.ms.model`.
```{r postcheck,eval = TRUE,fig.height = 6, fig.width = 6}
plot(
  coda::as.mcmc(out$post.samp.ms.model)
)
```
  
  
The posterior samples for all hypotheses are available in a list.
```{r postsamp,eval = TRUE,fig.height = 6, fig.width = 6}
names(
   out$post.samp
)

#For each of these list is a list of chains
length(out$post.samp[[1]])

#Here are the means of the posterior samples of all hypotheses for chain 1
lapply(
  out$post.samp,FUN = function(x){colMeans(x[[1]])}
)
```


# Plotting

Using the packages bayesplot and ggplot2, we can examine our posterior distributions along with the true probabilities values used to simulate
the data for this example.

```{r plot, eval = TRUE,fig.height = 4, fig.width = 6}
library(ggplot2)
library(bayesplot)

posteriors <- coda::as.mcmc(
  out$post.samp.ms.model
)

plot_title <- ggplot2::ggtitle(
  "Posterior distributions",
  "with medians and 80% intervals"
)
bayesplot::mcmc_areas(
  posteriors, prob = 0.8
  ) + 
  plot_title + 
  ggplot2::geom_vline(
    xintercept = p[1],
    linetype="dashed",
    color = c("red"),
    size = 1
  ) +
  ggplot2::geom_vline(
    xintercept = p[2],
    linetype="dashed",
    color = c("purple"),
    size = 1
  ) +
  ggplot2::geom_vline(
    xintercept = p[3],
    linetype="dashed",
    color = c("green"),
    size = 1
  )


```

Another way to examine our hypothesis is to plot the theoretical niche space for the hypothesis along with the posterior samples from the most supported model. To add the posterior samples you just need to provide them to the `posteriors` argument. First, let's plot the posterior samples out on the most supported model

```{r 3dplot 1,fig.align = "center",fig.height = 6, fig.width = 6,out.width = "6.5in"}
triplot(
  hyp = out$ms.model,
  more.points = TRUE,
  posteriors = out$post.samp.ms.model
)
```

Or we can plot the posteriors out on the whole hypothesis set. Unlike the previous plot this is done by default if you supply the fitted diel list object.

```{r 3dplot 2,fig.align = "center",fig.height = 6, fig.width = 6,out.width = "6.5in"}
triplot(out)
```


# The Unconstrained Model

If you are interested in comparing the constrained hypotheses to an unconstrained model, you can do that by adding it into the hypothesis set

```{r uncond,fig.align = "center",fig.height = 6, fig.width = 6,out.width = "6.5in"}
hyp.set.new <- c(
  hyp.sets("General"),
  "Uncon"
)
out <- diel.fit(
  y,
  hyp.set = hyp.set.new,
  n.chains = 2,
  post.fit = FALSE
)

#Compare the uncosntrained model to the rest of the models of the General hypothesis set
out$bf.table

```


# Data availabilty and an unequal prior

The package includes a set of available data that can be accessed as 

```{r avail.data}
data(diel.data)
head(diel.data)
```

For an explanation of each column in this dataset, check out the help file (i.e., `?diel.data`). For this example, we will use one of the data sets to fit a model with 1) equal weight on each hypotheses and 2) prior weight that is higher on the literature definition. 

```{r prior}
# Virginia Opossum data from the Urban Wildlife Information Network
dat <- diel.data[2,]

dat

# get just the data we need to fit the model. These must be in the 
#  same order specified here.
y <- data.frame(
  twilight = dat$twilight,
  day = dat$day, 
  night = dat$night
)
# provide species name to the row
rownames(y) <- dat$Common_name

y

# Model Comparison using equal eights
out1 <- diel.fit(
  as.matrix(y),
  hyp.set = hyp.sets("Traditional"),
  post.fit = FALSE
)

# Model Comparison using 0.8 probability weigh on the nocturnal hypothesis
out2 <- diel.fit(
  as.matrix(y),
  hyp.set = hyp.sets("Traditional"),
  post.fit = FALSE,
  prior = c(0.2/3,0.8,0.2/3,0.2/3)
)

# Results out1
round(
  out1$bf.table,
  digits = 2
)

# Results out 2
round(
  out2$bf.table,
  digits = 2
)
```
We see that the model weight on the nocturnal hypothesis increase because of our prior weight. 


Let's consider another case where we simulate 10 data sets with differing sample sizes to see how the prior weight affects the model weights. Specifically, we will put a high weight (0.95) on the Cathemeral Traditional hypothesis while the simulated data is from the Nocturnal hypothesis.

```{r prior2,  eval = TRUE, fig.align = "center",fig.height = 5, fig.width = 5}
set.seed(451)

n <- matrix(
  seq(10, 200, 20)
)

sim.dat <- t(
  apply(
    n,
    1,
    FUN = function(x){
      sim.diel(n.sim = 1, n.sample=x, hyp = "N")$y
    }
  )
)


prior.func <- function(x,prior){
  out = diel.fit(
    t(
      as.matrix(x)
    ),
    hyp.set=hyp.sets("Traditional"),
    post.fit = FALSE,
    print = FALSE,
    prior = prior
  )
  # get the probability of support for each hypothesis
  out$bf.table[,2]
}

#Equal priors
results.prior1 <- apply(
  sim.dat,
  1,
  prior = c(0.25,0.25,0.25,0.25),
  FUN = prior.func
)

# Unequal priors
results.prior2 <- apply(
  sim.dat,
  1,
  prior = c(0.05/3,0.05/3,0.05/3,0.95),
  FUN = prior.func
)

# do plot
matplot(
  t(results.prior1),
  type = "l",
  lwd = 3,
  ylab = "Model Probability",
  xaxt = "n"
)

# add x axis
axis(
  1,
  at = 1:length(n),
  lab = n
)

# add legend
legend(
  "right",
  legend = rownames(results.prior2),
  col = 1:4,
  lwd = 3,
  lty = 1:4
)

# do the same for unequal priors
matplot(
  t(results.prior2),
  type = "l",
  lwd = 3,
  ylab = "Model Probability",
  xaxt = "n"
)
axis(
  1,
  at = 1:length(n),
  lab = n
)
legend(
  "right",
  legend = rownames(results.prior2),
  col = 1:4,
  lwd = 3,
  lty = 1:4
)
```

The differences in the model probabilities are entirely due to the prior weights on each hypothesis. In the second plot, the likelihood of the data is trying to overwhelm such a highly certain prior.


# Plotting Issues in RStudio

Plotting is done using the package `plotly`, which can have issues with RStudio. If you are using RStudio and no figures are opening then: Tools--\>Global Options--\>Advanced--\>Rendering Engine Choose "Desktop OpenGL{} and then restart RStudio.

