---
title: "Example 5"
output: rmarkdown::html_vignette
always_allow_html: true
vignette: >
  %\VignetteIndexEntry{Example 5}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{Diel.Niche, suncalc, lubridate}
---

# Diel Selection
#### Author: Mason Fidino, Aka Dr. Rockstar
#### Date: 2023-06-15

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```


We will consider the hypothesis set, *Selection*, to make inference on diel periods that are used more than the amount of time available. 


# The Setup

```{r load}
# Load packages
  library(Diel.Niche)
  library(lubridate)
  library(suncalc)
```

We will use data available in the package (`diel.data`) provided by the Urban Wildlife Information Network (https://www.urbanwildlifeinfo.org/), coordinated by the Lincoln Park Zoo's Urban Wildlife Institute . The specific data are camera trap detections of the white-tailed deer (*Odocoileus virginianu*) collected in Chicago, Illinois during 2020. Since there are 5 different data
sets from 2020, we are only going to grab one dataset from the summer and another from the fall.


```{r data}
data(diel.data)

# Define a year variable
  diel.data$min_year  <-lubridate::year(
    lubridate::mdy(
      diel.data$min_date
    )
  )

# Extract data
  deer <- subset(
    diel.data,
    scientificName == "Odocoileus virginianus" &
    min_year == "2020"
  )


# subset to first two
  deer <- deer[deer$min_date %in% c("7/5/2020", "10/2/2020"),]

# Visualize the data
  head(deer)
```

One small complication that makes the selection hypothesis set different from
the other hypothesis sets is that it requires you to input the proportion of
time each diel period takes up during a given time of year. This can be done using the `suncalc` package in `R` via `getSunlightTimes()`. This function allows a vector of dates but only a single latitude and longitude. Fortunately, `diel.data` already has the average location of Chicago, Illinois. For your own work, you could likely just take the average of all your camera traps in a given
study area.

```{r suncalc}
  chicago.hours <- suncalc::getSunlightTimes(
    date = lubridate::mdy(deer$min_date),
    lat = unique(deer$mean_lat),
    lon =  unique(deer$mean_lon),
    keep = c("dusk", "night", "dawn","nightEnd"),
    tz = "US/Central"
  )

# calculate the hours in each diel period
twilight.hours <- (chicago.hours$dawn - chicago.hours$nightEnd) +
  (chicago.hours$night - chicago.hours$dusk)
day.hours <- chicago.hours$dusk - chicago.hours$dawn
night.hours <- 24 - (twilight.hours + day.hours)

# Add some new columns to diel.data, but also make these hours
#  into proportions.
deer$prop_twilight <- as.numeric(twilight.hours / 24)
deer$prop_day <-  as.numeric(day.hours / 24)
deer$prop_night <- as.numeric(night.hours / 24)
```

Now that we have added all of these proportions, we just need call `diel.fit()`
to estimate selection. However, in order to do this we need to modify the
inequalities used to fit this model with `diel.ineq()`. And while we calculated
all three proportions (twilight, day, and night), we only need to provide the 
first two to `diel.ineq()`. As there are two datasets from 2021, let's fit
each individually and inspect the results. But since we have data from different
parts of the year, let's also plot out the visual representation of this diel
hypothesis set to see how they differ.

For this first dataset, the `triplot` looks like


```{r first_triplot,fig.height = 6, fig.width = 6}

# first dataset
  selection.ineq1 <- diel.ineq(
    p.avail =  c(
      deer$prop_twilight[1],
      deer$prop_day[1]
    )
  )

  triplot(
    hyp = hyp.sets("Selection"),
    diel.setup = selection.ineq1
  )

```

Notice that the largest parameter spaces are defined by the `Night-Twilight`, `Night`, and `Twilight` hypotheses. This is because twilight and then night have lower available time and thus more combinations of parameters in which use can be greater than availability.

The second `triplot` shows the changing availability reduces the `Night-Twilight` hypothesis and increases ore `Twilight` hypothesis:

```{r triplot2,fig.height = 6, fig.width = 6}
  selection.ineq2 <- diel.ineq(
    p.avail =  c(
      deer$prop_twilight[2],
      deer$prop_day[2]
    )
  )

  triplot(
    hyp = hyp.sets("Selection"),
    diel.setup = selection.ineq2
  )

```


And now, to the analysis. For the first dataset we could use the already
created `selection.ineq1`
{
```{r analysis1}

# subset data to response variables
  deer.y1 <- as.matrix(
    deer[1,c("twilight", "day", "night")]
  )

  out1 <- diel.fit(
    deer.y1,
    hyp.set = hyp.sets("Selection"),
    diel.setup = selection.ineq1,
    n.chains = 2,
    post.fit = TRUE
  )
  
  out1$bf.table
  
```
The most supported model for this dataset (`N.avail`) indicated that white-tailed deer are using the night diel period far more than it is available. Looking at the raw data (i.e., `deer.y1`), about 45% of the camera images collected occurred at night. However, night only accounted for 19% of the diel time during a day.

```{r}

round(
  deer.y1 / sum(deer.y1),
  2
)

round(
  deer[1,c("prop_twilight", "prop_day", "prop_night")],
  2
)

```

If we wanted to look at a histogram that quantifies the selection for a given
diel period, we can do so by taking the MCMC output of the best fit model and dividing that respective category by the proportion of time it was available. When a species is selecting for a diel period more than it is available, then the posterior distribution should be above 1, which in this case it clearly is.

```{r avail_plot1,fig.height = 6, fig.width = 6}

hist(
  out1$post.samp.ms.model[,"p_night_1"] / deer$prop_night[1],
  main = "",
  xlim = c(0,3),
  las = 1,
  xlab = "Selection for nighttime activity relative to availability"
)
abline(v = 1, lty = 2)

```

And finally, we can plot the posterior out ontop of the full parameter space for the Selection hypothesis set with `triplot()`.

```{r triplot_posterior1,fig.height = 6, fig.width = 6}

triplot(
  out1,
  diel.setup = selection.ineq1
)

```

And here is the same analysis, but with the second dataset from the fall, where we find slightly different results!

```{r dataset2analysis}

# subset data to response variables
  deer.y2 <- as.matrix(
    deer[2,c("twilight", "day", "night")]
  )

  out2 <- diel.fit(
    deer.y2,
    hyp.set = hyp.sets("Selection"),
    diel.setup = selection.ineq2,
    n.chains = 2,
    post.fit = TRUE
  )
  
  round(
    out2$bf.table,
    2
  )

```

For this analysis, Day-Night Selection (`D.N.avail`) was the most supported hypothesis, though support for it was much weaker with a model probability of 0.71. For this dataset, it appears that deer spent most of their time active during either the day or night, which together represented about 91% of a given day. 


```{r}

round(
  deer.y2 / sum(deer.y2),
  2
)

round(
  deer[2,c("prop_twilight", "prop_day", "prop_night")],
  2
)

```


For plotting out selection, we need to do so separately for each diel period as, 

```{r avail_plot2,fig.height = 6, fig.width = 6}
hist(
     out2$post.samp.ms.model[,c("p_day_1")]/deer$prop_day[2],
     freq = FALSE,
     xlim=c(0.8,1.5),
     main="",
     col="yellow",
     las=1,
     xlab = "Selection for day and nighttime activity relative to availability"
)

hist(
     out2$post.samp.ms.model[,c("p_night_1")]/deer$prop_night[2],
     freq = FALSE,
     col="blue",
     add=TRUE
)
```

The first noticeable thing is the hard constraint at 1, truncating the posterior distributions. This makes sense why the model probability support was not that high. The posteriors are near 1 and relatively variable. However, given this model, both day and night are to a small degree being selected for more than available. However, this selection may be 'biologically significant' when placed in the larger context of their overall behavior and life history pattern. Plotting this out on via `triplot()` you can see that the entire `Day-Night (Selection)` parameter space is filled with the posterior.

```{r triplot_posterior2,fig.height = 6, fig.width = 6}

triplot(
  out2,
  diel.setup = selection.ineq2
)

```

# Conclusion

The Selection hypothesis set is unlike all the other flavors of hypotheses within `Diel.Niche` because it can be used to quantify use relative to availability in a way that is similar to resource selection functions. This additional nuance allows users of `Diel.Niche` to not only classify the diel phenotype of a given species, but also to determine if there are specific diel periods that are used more than they are available. This could be of particular interest for species that are classified as Crepuscular, given that the available time in this diel period will usually be relatively small. While White-tailed deer are considered crepuscular, our analysis showed that they actually used night much more than other diel periods, which may be the result of deer modifying their activity levels throughout urban Chicago. Conversely, deer may be selecting for night during the summer given how hot it can get throughout Chicago as a way to better regulate their body temperature.