---
title: "Example 2"
output: rmarkdown::html_vignette
always_allow_html: true
vignette: >
  %\VignetteIndexEntry{Example 2}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{ggplot2,ggthemes,lubridate,Diel.Niche}
---

# Multiple species with one data analysis unit
#### Author: Brian D. Gerber
#### Date: 2023-05-10

```{r start, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Commonly, researchers are interested in making inference on the spatial and temporal activity of an entire meso/large mammal community. Here, we will consider a camera trap study that is aimed at sampling multiple animal species in an area to make inference on the diel niche of each species. We will use data available in the package (`diel.data`) provided by the Urban Wildlife Information Network (https://www.urbanwildlifeinfo.org/), coordinated by the Lincoln Park Zoo's Urban Wildlife Institute. The specific data are camera trap detections of the urban mammal community during the winter of 2019 in Chicago, Illinois USA. The available data are aggregated independent counts from 131 camera locations. Our objective is to evaluate the support for the Gen earl diel hypotheses and compare the diel niche support of each species, as well as compare these results with literature designations of diel activity. 

# The Setup

```{r setup}
# Load package
  library(Diel.Niche)
  library(lubridate)
  library(bayesplot)
  library(ggplot2)
  library(ggthemes)

  data("diel.data")
# Define year variable
  diel.data$min_year <- lubridate::year(
    as.POSIXct(
      diel.data$min_date,
      format = "%m/%d/%Y"
    )
  )

# Extract winter data in 2019
  winter <- subset(
    diel.data,
    season=="Winter" & min_year=="2019"
  )

# Data visual
  head(winter)

# Species observed
  unique(winter$scientificName)

```


We can extract all the species data into object `y` as,

```{r data}
y  <-  winter[,c("twilight", "day", "night")]

rownames(y)=winter$Common_name

y
```

# Model Comparison

To get model probabilities for each species, we can define a new function that uses `diel.fit()` and pass our data `y` to the new function using `apply()`.

```{r Analysis}
multi.fit.fun <- function(y){
    out = diel.fit(
      t(
        as.matrix(y)
      ),
      hyp.set = hyp.sets("Traditional"),
      post.fit = FALSE, 
      prints=FALSE
    )
  # Get the Bayes factors for each hypothesis 
  list(
    ms.model = out$ms.model,
    prob = out$bf.table
  )
}

out.multi <- apply(
  y,
  1,
  multi.fit.fun
)
```

Next, we need to extract from our list the results of interest, including the model results table for each species and the most supported model and its probability. 

```{r results}
# The most supported hypothesis for each species and its probability

#The probability set for each species
  temp <- sapply(
    out.multi,
    function(x) x["prob"]
  )

  sp.model.probs <- matrix(
    unlist(
      lapply(
        temp,
        function(x) x[,"Posterior"]
      )
    ),
    ncol = length(temp),
    byrow = TRUE
  )
  # transpose so it is species on row, models on columns
  sp.model.probs <- t(sp.model.probs)
  
  # give names that make sense
  rownames(sp.model.probs) <- rownames(y)
  colnames(sp.model.probs) <- rownames(temp[[1]])

  round(sp.model.probs,digits=2)

# Get the codes of the most supported hypothesis for each species.
  ms.hyps <- unlist(
    lapply(
      out.multi,'[',1)
  )
# give species names to each element
  names(ms.hyps) <- names(out.multi)
  ms.hyps

# and the probability for that hypothsis given the data.
  prob.hyps <- unlist(
    lapply(
      lapply(
        out.multi,
        '[',2
      ),
      FUN=function(x){max(x$prob[,"Posterior"])}
      )
  )

# Combine both into a data.frame
  diel.niche.df <- data.frame(
    hypothesis = ms.hyps,
    probability =round(prob.hyps, 2)
  )
  diel.niche.df
```

We find that there is clear evidence (model probability at/near 1.0) that White-tailed deer are cathemeral (Traditional), Northern Raccoon and Eastern Cottontail are nocturnal, and the two squirrel species are diurnal. We are less confident with model probabilities between 0.6 and 0.7 that Coyote are cathemeral (Traditional) and Virginia Opossum are nocturnal. 

Comparing these results to the literature (Wilson et al. 2001-2019), we see agreement with regard to the Coyote, Virginia Opossum, Northern Raccoon, and both squirrel species. There is however disagreement regarding the White-tailed Deer and the Eastern cottontail, which are considered crepuscular. 

The reason why there is likely a departure from what is reported in the literature and these two crepuscular species is that so far there have not been quantitative definitions for different diel phenotypes. Twilight is the shorted diel period, and for these species to be classified as Crepuscular they would have to spend at least 80% of  their time active within a small window. As these species have not done this, and have instead used multiple diel periods, the Traditional hypothesis set 
classified them as cathemeral. Instead, it is more likely that 'classically' crepuscular species are using the twilight period more than it is available (i.e., these species are not constraining their activity, but rather selecting for a given
diel period). This could be quantified with the Selection hypothesis set, and for an example of this see the `Example5` vignette in `Diel.Niche`.

# Model Parameters 

Let's now fit each species' most supported models to get posterior distributions for probabilities of activity. To do so, we make a new function that executes `diel.fit()`, but this time specifying `post.fit = TRUE` to get the posterior samples.


```{r Analysis2}

y.df <- data.frame(
  y,
  hyp=diel.niche.df$hypothesis
)

multi.fit.fun2 <- function(y.df){
    out = diel.fit(
      t(
        as.integer(
          y.df[1:3]
        )
      ),
      hyp.set = y.df[4],
      post.fit = TRUE,
      prints=FALSE, 
      n.chains = 3,
      n.mcmc = 2000,
      burnin = 1000
  )

  list(
    post.samp = out$post.samp[[1]],
    gelman.diag=out$gelm.diag
  )
}

out.multi2 <- apply(
  y.df,
  1,
  multi.fit.fun2
)

# We can extract the geman-rubin diagnostics to check for convergence issues
  sapply(
    out.multi2,
    function(x) x["gelman.diag"]
  )
  
# Seeing no convergence issues, we can combine our chains
  post.samples <- lapply(
    sapply(
      out.multi2, 
      function(x) { x["post.samp"] }
    ),
    FUN = function(x) { do.call("rbind", x) } 
  )
  
```



```{r parameters}
# Summarize posterior model probabilities using quantiles

#all species, posterior quantiles for three parmaeters
  prob.quantiles <- lapply(
    post.samples,
    FUN = function(x) { 
      apply(
        x,
        2,
        quantile,
        probs = c(0.025, 0.5, 0.975)
      )
    }
  )

# as one example, here are the summarised samples for coyote
  prob.quantiles$Coyote.post.samp

# Extract posterior medians fo each species
  prob.median <- matrix(
    unlist(
      lapply(
        prob.quantiles,
        FUN = function(x) { x[2,] }
      )
    ),
    ncol = 3,
    byrow = TRUE
  )
  rownames(prob.median)=row.names(y)
  colnames(prob.median)=colnames(prob.quantiles$Coyote.post.samp)
  prob.median
```

# Plotting

Lets create a straightforward plot of the probabilities of activity for each species during the three diel periods.

```{r plotting, eval = TRUE,fig.height = 6, fig.width = 8}

  post <- post.samples
  n.species <- length(post)
  
# get the posterior for each species
  post <- lapply(
    post,
    FUN = function(x) {
      colnames(x) = c("P(twilight)","P(daytime)","P(nighttime)");
      x
    }
  )
# bind each MCMC chain together for each species
  post <- do.call(
    'rbind',
    lapply(
      post,
      FUN = function(x) {
        bayesplot::mcmc_intervals_data(
          x,
          prob = 0.5,
          prob_outer = 0.95
        )
      }
    )
  )
# add species names to each of their three probabilities
  post$Species <- rep(
    rownames(y),
    each = 3
  )

# position of y for plotting purposes for each species,
#  jittering each a tiny bit.
  y.pos <- rep(
    rnorm(
      nrow(post)/3,
      0,
      0.07
    ),
    each = 3
  )
  
  pos <- position_nudge(
    y = y.pos,
    x = rep(0,nrow(post))
  )
  
  p <- ggplot(
    post,
    aes(x = m, y = parameter, color = Species)
  ) + 
    geom_point(
      size = 6,
      position = pos
    ) +
    geom_linerange(
      aes(xmin = ll, xmax = hh),
      position = pos
    ) +
    xlab("Probabilty") +
    ylab("") +
    xlim(0,1) +
    theme(text = element_text(size = 20)) + 
    coord_flip()+
    scale_colour_colorblind()+
    theme(
      panel.background = element_rect(fill='white', colour='black'),
      legend.key=element_rect(fill="white")
    )
      
    p      

```

We see that the the two species most active during twilight are the Eastern Cottontail and the Coyote. The Fox and Gray Squirrels are almost only active during the daytime, thus confirming why they were classified as diurnal. The white-tailed deer is most active during the day but still about a 0.27 probability of being active at nighttime. The Eastern Cottontail, Virginia Opossum, and the Northern Raccoon are highly active during the nighttime with probabilities greater than 0.80. 

# References

Wilson, D.E. Rylands, A.B., Lacher, T.E., Mittermeier, R.A. (2001-2019) Handbook of the mammals of the world - Volume 1-3, 5-9, (Lynx Edicions, Barcelona)