---
title: "Example 6"
output: rmarkdown::html_vignette
always_allow_html: true
vignette: >
  %\VignetteIndexEntry{Example 6}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
  %\VignetteDepends{Diel.Niche, suncalc, lubridate, momentuHMM, sf, ggplot2, ggpubr,conicfit}
---

# Diel Activity from State-Dependent Animal Movement  
#### Author: Brian D. Gerber
#### Date: 2023-09-12

```{r setup, include=FALSE, eval=TRUE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)

#save(crwOut, m1, m2, m3, file="Example6.RData")

load("Example6.RData")

```

In this example We consider how to use an animal movement behavior analysis to provide inference on diel activity using the $\texttt{Diel.Niche}$ R package. This is by no means the only way to think about animal movement and diel activity. Specifically, we use the R package $\texttt{momentuHMM}$ to fit a hidden Markov model that assumes animal location data is a bivariate time series to estimate step lengths and turning angle's at each time point. We will consider state-dependent movement, where we estimate the behavioral state (e.g., foraging or traveling) of each location based on step lengths and turning angles. We also consider diel cycles of temperature as a covariate when modeling movement parameters. 


# The Setup

```{r load, message=FALSE}
# Load packages
  library(Diel.Niche)
  library(lubridate)
  library(suncalc)
  library(momentuHMM)
  library(sf)
  library(ggplot2)
  library(ggpubr)
```

We will use African Elephant (*Loxodonta africana*) data available from Wall et al. 2014 and published online at [MoveBank](http://dx.doi.org/10.5441/001/1.f321pf80). 

The animal movement script that fits the hidden Markov model is adapted from the example outlined in $\texttt{momentuHMM}$ and available on GitHub [https://github.com/bmcclintock/momentuHMM/blob/master/vignettes/examples/elephantExample.R](https://github.com/bmcclintock/momentuHMM/blob/master/vignettes/examples/elephantExample.R).

# Movement Data and Iniital Setup

First, we need to load the data and setup the data structure for animal movement modeling. 

```{r data, eval = TRUE, echo = TRUE}
# Read the data locally - NEEDS TO BE CHANGED
  rawData<- read.csv(
    "Wall et al. 2014 African Elephant Dataset (Source-Save the Elephants).csv"
  )
```

```{r Movement Model Setup from momentuHMM, eval = TRUE, echo = TRUE}

# Select and rename relevant columns
  rawData <- rawData[,
    c(
      "individual.local.identifier",
      "timestamp",
      "location.long",
      "location.lat",
      "external.temperature"
    )
  ]
  colnames(rawData) <- c("ID","time","lon","lat","temp")

# There are two individual's tracks. Only keep the  first track - named, "Salif Keita"
  rawData <- subset(
    rawData,
    ID==unique(ID)[1]
  )

# Convert times from factors to POSIX
  rawData$time <- as.POSIXct(
    rawData$time,
    tz="GMT"
  )

  
# Project to UTM coordinates using sf
  llcoord <- sf::st_as_sf(
    rawData[,3:4],
    coords = c("lon","lat"),
    crs="+proj=longlat +datum=WGS84"
  )
  utmcoord <- sf::st_transform(
    llcoord,
    crs="+proj=utm +zone=30 ellps=WGS84"
  )

# Add UTM locations to data frame
  rawData$x <-  sf::st_coordinates(utmcoord)[,1]
  rawData$y <-  sf::st_coordinates(utmcoord)[,2]
```

Now with the data all setup, we can standardize the location intervals at hour intervals by using the function `crawlWrap` which implements a continuous-time correlated random walk model.


```{r crawl, eval = FALSE, echo = TRUE}
# Model and Predict locations using crawl - accomadates missing data at one hour intervals
  crwOut <- momentuHMM::crawlWrap(
    rawData,
    timeStep="hour",
    theta=c(6.855, -0.007),
    fixPar=c(NA,NA)
  )
```


Using the output, we can plot the observed and predicted standardized locations.

```{r movement plot1, fig.width = 6, fig.height = 6, eval = TRUE, echo = TRUE}
  plot(
    crwOut,
    ask = FALSE
  )
```

# Fitting hidden Markov Models

Now that we have standardized fixed-location intervals, we can estimate state-dependent turning angles and step lengths. First, we will setup our data for model fitting in `momentuHMM`. Overall, we are going to fit three models to these data. The first is a null model (`m1`), the second incorporates some covariates on transition probabilities (`m2`), and the third incorporates covariates onto multiple parameters (`m3`; i.e., transition probabilities, variation in step length, and turning angle concentration).

```{r HMM1 setup, eval = TRUE, echo = TRUE}
# Create momentuHMMData object from crwData object
  elephantData <- momentuHMM::prepData(
    data=crwOut,
    covNames="temp"
  )

# Add cosinor covariate based on hour of day
  elephantData$hour <- as.integer(
    strftime(
      elephantData$time,
      format = "%H",
      tz="GMT"
      )
  )

# Label states
  stateNames <- c("encamped","exploratory")

# Distributions for observation processes
  dist = list(step = "gamma", angle = "wrpcauchy")

# Initial parameters
  Par0_m1 <- list(
    step=c(100,500,100,200),
    angle=c(0.3,0.7)
  )
```

The code to fit the null model is... 

```{r fit HMM1, eval = FALSE, echo = TRUE}
# Fit hidden markov model with 2 states
  m1 <- momentuHMM::fitHMM(
    data = elephantData,
    nbStates = 2,
    dist = dist,
    Par0 = Par0_m1, 
    estAngleMean = list(angle=FALSE),
    stateNames = stateNames
  )
```

Next, let's consider a model (`m2`) where the transition probabilities between states vary by temperature and the interaction with a cyclical function based on the hour of the day. See [https://cran.r-project.org/web/packages/momentuHMM/vignettes/momentuHMM.pdf](momentuHMM vignette) for more details. 

```{r HMM2 setup, eval = TRUE, echo = TRUE}
# Formula for transition probabilities
  formula <- ~ temp * cosinor(hour, period = 24)

# Initial parameters (obtained from nested model m1)
  Par0_m2 <- momentuHMM::getPar0(
    model=m1,
    formula=formula
  )
  
  # And fit the model
  m2 <- momentuHMM::fitHMM(
    data = elephantData,
    nbStates = 2,
    dist = dist,
    Par0 = Par0_m2$Par, 
    beta0=Par0_m2$beta,
    stateNames = stateNames,
    formula=formula
  )

```

Lastly, we will consider a more complex model (m3) where the transition probabilities and mean and standard deviation of step lengths vary by the cyclicality of temperature, and the concentration of the turning angle varies by temperature. See [https://cran.r-project.org/web/packages/momentuHMM/vignettes/momentuHMM.pdf](momentuHMM vignette) for more details on using cyclical or spline functions to model complex non-linear variation. 

```{r HMM3 setup, eval = TRUE, echo = TRUE}
# Formulas for parameters of state-dependent observation distributions
  DM <- list(
    step = list(
      mean = ~ temp * cosinor(hour, period = 24),
      sd = ~ temp * cosinor(hour, period = 24)
    ),
    angle = list(
      concentration = ~ temp
    )
  )

# Initial parameters (obtained from nested model m2)
  Par0_m3 <- momentuHMM::getPar0(
    model=m2,
    formula=formula, DM=DM
  )

# And fit the model
  m3 <- momentuHMM::fitHMM(
    data = elephantData,
    nbStates = 2,
    dist = dist,
    Par0 = Par0_m3$Par, 
    beta0 = Par0_m3$beta,
    DM = DM,
    stateNames = stateNames,
    formula = formula
  )
```

Now we can fit m3....

```{r fit HMM3, eval = FALSE, echo = TRUE}
# fit model 3

```

We can compare the three models using Akaike's Information Criterion (AIC). 
```{r HMM model commparison, eval = TRUE, echo = TRUE}
# Calculate and compare AIC
  AIC(m1,m2,m3)
```

Using AIC as an estimate of the Expected Kullback-Leibler discrepancy, we see that the most complex model, model 3, is the most supported. This is because it is has the lowest AIC value, which means it is the model with the minimum difference from the theoretical generating model, relative to the other two models in the model set. 

```{r HMM3 parameters, eval = TRUE, echo = TRUE}
# Show model parameters
  m3
```



There are a lot of results to consider from this model. We encourage those interested to follow up with the [https://cran.r-project.org/web/packages/momentuHMM/vignettes/momentuHMM.pdf](momentuHMM vignette). However, notice that the mean step-lengths are much different between the `encamped` (185 m) and `exploratory` (866 m) movement states. Further, looking at the transition probabilities between states tells us that when this individual is moving according to either of the two behavioral states, they are more likely to stay in the same state between 1 hour location intervals (at mean covariate values). 

Now lets plot the covariate effects and movement track along with the estimated movement states.

```{r plot HMM, fig.width = 6, fig.height = 10, eval = TRUE, echo = TRUE}
  plot(m3, plotCI = TRUE, covs = data.frame(hour=12))
```

Some interesting take-aways,

* The mean and standard deviation of step lengths decrease in the `encamped` state as temperature increases
* The mean and standard deviation of step lengths at the mean temperature vary cyclically over the diel period
* The mean and standard deviation of step lengths decrease very little or not at all in the `exploratory` state as temperature increases 
* Turning angle concentratoin decreases as temperatures increase during the `encamped` state and less so in the `exploratory` state.
* The probability of staying in the `encamped` state increases as temperatures increase.
* There appears to be moderate cyclicality over the diel cycle in transitioning between states

Before leaving the hidden Markov model, we need to derive the states associated with each location. 

```{r HMM states, eval = TRUE, echo = TRUE}
# Decode most likely state sequence
  states <- viterbi(m3)
  states[which(states==1)] <- "encamped"
  states[which(states==2)] <- "exploratory"

# Derive percentage of time spent in each state
  table(states)/nrow(elephantData)
```

# Diel Activity 

Let's turn to the $\texttt{Diel.Niche}$ R package and estimate the diel phenotype of this individual by behavioral state. Since the individual was tracked continuously from March 2008 to September 2010, lets consider their diel phenotype on a rolling monthly basis. To do so, we need to classify each location time into the appropriate twilight, daytime, and nocturnal bins.

```{R Diel.Niche Data Setup, eval = TRUE, echo = TRUE}
#Get time period cutoffs for each date/time  using the package suncalc 
  tod <- suncalc::getSunlightTimes(
    date = date(date(elephantData$time)),
    lat = mean(elephantData$lat,na.rm=TRUE),
    lon =  mean(elephantData$lon,na.rm=TRUE),
    keep = c("dusk", "night", "dawn","nightEnd"),
    tz = "GMT"
  )  

# Take these and extract the hours, minutes, and seconds    
  time.subset=data.frame(apply(tod[,4:7],2,strftime,format="%H:%M:%S"))

# Get Elephant times in hours, minutes, seconds
  elephant.times=strftime(elephantData$time, format="%H:%M:%S")

# These should be the same length  
  length(elephant.times)==nrow(time.subset)
  
# Create a vector to store how each observation is defined- twilight, daytime, and nighttime.
  time.cat=rep("nighttime",nrow(time.subset))

# Find times that are b/w dawn and dusk and define that as daytime  
  index.day=which(elephant.times>time.subset$dawn & elephant.times<time.subset$dusk)

# Find times that are before dawn (begining of daytime) and after the nightEnds
  index.dawn=which(elephant.times<time.subset$dawn & elephant.times>time.subset$nightEnd)
  
# Find times that are after dusk and before the begining of night
  index.dusk=which(elephant.times>time.subset$dusk & elephant.times<time.subset$night)

# Change vector elements correspond to daytime and twilight  
  time.cat[index.day]="daytime"  
  time.cat[c(index.dawn,index.dusk)]="twilight"  

# For the entire smapling period this is the frequency of classifications, disregarding particular sampling periods and behavioral state
  table(time.cat)
  
# Create a new data.frame of dates, states, and time category and then split data into a list of elements
# Month and Year of sampling
  Month.Year <- paste(month(elephantData$time),"/",year(elephantData$time),sep="")

  time.cat.dates=data.frame(Month.Year,states,time.cat)

  dat.list <- split( time.cat.dates$time.cat , f = as.factor(paste(time.cat.dates$Month.Year,time.cat.dates$states, sep="-")))
  
# Here is the sample size per month and behavioral state of sampling period  
  n.obs <-  as.vector(unlist(lapply(dat.list,FUN=length)))
  n.obs

# The General hypothesis set requires a fairly reasonable sample size. Let's drop sampling periods with < 50 observations.
  dat.list=dat.list[-which(n.obs<50)]
  
  
# Create a frequency table  and arrange columns in the order: Twilight, Daytime, and Nighttime 
  dat.table <- lapply(dat.list,FUN=function(x)table(x)[c(3,1,2)])
  n.units <- length(dat.table)

# Check to make sure there are there are no missing values
  unlist(lapply(dat.table, FUN=function(x){which(is.na(x))}))

```


We have the data that we need in the object `dat.table`. Each list elements includes the frequencies for twilight, daytime, and nighttime for a specific movement behavioral state and month of the year.

Next, we need to use the `diel.fit` function to estimate the most probable hypothesis within the General hypothesis set. Using the `lapply` function we can fit each dataset quickly.

```{r Diel.Niche model fitting, eval = TRUE, echo = TRUE}
# For each list element, fit the General hypothesis set and extract the most probable hypothesis
    elephant.hyps <- lapply(dat.table, FUN=function(x){
                           fit = diel.fit(t(matrix(x)),
                                          hyp.set = hyp.sets("General"),
                                          post.fit=FALSE,
                                          prints = FALSE
                                         )
                            fit$ms.model
})
  

#  Grab the state names
  states.by.hyp = sub('.*-', '', as.character(names(elephant.hyps)))
  
# Frequency table of Supported Hypotheses by Behavioral States where each sampling period is 1 month from March 2008 to September 2010. 
  table(states.by.hyp,unlist(elephant.hyps))

# This individual is always following a Diurnal-Nocturnal strategy. 
```

Now that we have the most supported model, we can use the `diel.fit` function to estimate the posterior distributions of activity for each sampling period and behavioral state. To change it up, lets use a for loop to loop through each data set, saving the posterior median and quantiles that will then use for plotting.

```{r, diel posteriors, eval = TRUE, echo = TRUE}

  # For each list element, get the posterior distributions from the most probable hypothesis. probable hypothesis

  crep = day = night =NULL
  for(i in 1:length(elephant.hyps)){
    out <-   diel.fit(t(matrix(dat.table[[i]])),
                     hyp.set = unlist(elephant.hyps)[i],
                     post.fit=TRUE,
                     bf.fit = FALSE,
                     prints = FALSE
                    )
    #print(i)

    temp=t(apply(out$post.samp[[1]][[1]],2,quantile, probs=c(0.025,0.5,0.975)))
    crep=rbind(crep,temp[1,])
    day=rbind(day,temp[2,])
    night=rbind(night,temp[3,])
  }
  

# get year and month order for plotting
   month <-  as.integer(sub("/.*", "", names(elephant.hyps)))
   month <- sprintf("%02d", month)
   year <- sub('.*/', '', as.character(names(elephant.hyps)))
   year <- as.integer(sub("-.*", "", year))
   
   Sampling <- as.Date(paste(rep("01",length(month)),month,year,sep="/"), format = "%d/%m/%Y")

# Arrange the data for plotting
  crep.plot <- data.frame(crep, states.by.hyp,Sampling)
  colnames(crep.plot) <- c("LCL","median", "UCL", "State", "Sampling") 
  crep.plot <- crep.plot[order(Sampling),]
  crep.plot$Sampling <- factor(crep.plot$Sampling, ordered = TRUE)
  
  day.plot <- data.frame(day, states.by.hyp,Sampling)
  colnames(day.plot) <- c("LCL","median", "UCL", "State", "Sampling") 
  day.plot <- day.plot[order(Sampling),]
  day.plot$Sampling <- factor(day.plot$Sampling, ordered = TRUE)

  night.plot <- data.frame(night, states.by.hyp,Sampling)
  colnames(night.plot) <- c("LCL","median", "UCL", "State", "Sampling") 
  night.plot <- night.plot[order(Sampling),]
  night.plot$Sampling <- factor(night.plot$Sampling, ordered = TRUE)

```

Lastly, lets plot the posterior distributions for each of the periods of crepuscular, daytime, and nighttime by behavioral state and sampling period.

```{r diel activitiy ploting, fig.width = 6, fig.height = 10, eval = TRUE, echo = TRUE}
col2 <- c("darkgrey","purple")

p1 <- ggplot(crep.plot, aes(x=Sampling, y=median,group=State,fill=State)) +
            theme_bw()+ 
            geom_line(linetype = "solid", linewidth=2) + 
            ylim(0,1)+
            geom_ribbon(aes(x=Sampling,ymin=LCL, ymax=UCL,group=State),alpha=0.4)+
            scale_fill_manual(values=col2)+
            labs(y = "P(Twilight)")+ 
            labs(x = "Sampling Month")+ 
            theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

p2 <- ggplot(day.plot, aes(x=Sampling, y=median,group=State,fill=State)) +
            theme_bw()+ 
            geom_line(linetype = "solid", linewidth=2) + 
            ylim(0,1)+
            geom_ribbon(aes(x=Sampling,ymin=LCL, ymax=UCL,group=State),alpha=0.4)+
            scale_fill_manual(values=col2)+
            labs(y = "P(Daytime)")+ 
            labs(x = "Sampling Month")+ 
            theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

p3 <- ggplot(night.plot, aes(x=Sampling, y=median,group=State,fill=State)) +
            theme_bw()+ 
            geom_line(linetype = "solid", linewidth=2) + 
            ylim(0,1)+
            geom_ribbon(aes(x=Sampling,ymin=LCL, ymax=UCL,group=State),alpha=0.4)+
            scale_fill_manual(values=col2)+
            labs(y = "P(Nightime)")+ 
            labs(x = "Sampling Month")+ 
            theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))

ggarrange(p1, p2, p3, ncol=1, nrow=3, widths = c(1,1,1))

```

# Conclusion

We found that the most supported animal movement model for this individual African Elephant was one that allowed two movement states where the transition probabilities, step lengths, and turning angle parameters varied by the diel cycle and temperature. In regard to diel activity, we found that over more than two years of continuous sampling that this individual exclusively followed a Diurnal-Nocturnal diel phenotype. In one sampling month (3/2008) there was high nocturnal activity when in the `exploratory` movement state (~0.78 probability), but this was not enough to be classified as nocturnal. We also only found fairly smaller differences in diel activity between the two movement states (`encamped` and `exploratory`) across the entire sampling period. Twilight activity was also similar between the two movement states and had the least variability across monthly sampling periods.

# References

Wall J, Wittemyer G, LeMay V, Douglas-Hamilton I, Klinkenberg B. 2014. Elliptical time-density model to estimate wildlife utilization distributions. Methods Ecology and Evolution, 5:780-790. https://doi.org/10.1111/2041-210X.12218

